---
title: "My Study Notes & Guides"
author: "Marcella Harris"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

```{r theme-setup, echo=FALSE, results='hide', message=FALSE}
library(bslib)
bs_global_theme(bs_theme(
  version = 4,
  bootswatch = "lux",
  primary = "#b38962",  # Optional custom
  code_font = font_google("Azeret Mono"),
  heading_font = font_google("Cinzel"),
  base_font = font_google("Gruppo")
))
```

# Welcome

<img src="https://via.placeholder.com/150" alt="Logo or Banner" style="float: right; margin-left: 20px; width: 150px;">

Welcome to **Study Notes** — a curated collection of learning resources and guides covering various topics. This site is organized using [**bookdown**](https://bookdown.org) and is hosted on GitHub Pages.

Whether you're here to review, revise, or explore something new, you'll find each section designed to be clear, concise, and practical.

---

## 📘 What's Inside

This collection includes:

- Key concepts & summaries
- Real-world examples
- Code snippets (where applicable)
- Diagrams & visuals
- Learning tips

> ✍️ *Each topic is organized into its own chapter (see left sidebar).*


---

## 🚀 Getting Started

[Start reading »](studyGuideD421.html){.btn .btn-primary}

---

## 🙌 Contribute or Feedback

Feel free to view or contribute on [GitHub](https://github.com/MarcellaHarr/study-guides-and-notes.git), or open an issue if you find something to improve.
