
```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = FALSE, echo = TRUE)
```
###### Discrete Mathematics and Its Applications (8th Edition)
<br>

```{r mathLink, include=FALSE}
# Mathematical Notations/Symbols for R-markdown
# https://rpruim.github.io/s341/S19/from-class/MathinRmd.html
```

<hr>

# Chapter 1The Foundations: Logic and Proofs
<hr>

### Exercise 1a {.tabset .tabset-fade}

::: {.nav nav-tabs}

## Step-1 {.nav-item}

### {.nav-link active}

In this exercise, we determine whether the given statement represents a proposition.

**What is a proposition?**



## Step-2 {.nav-item}

### {.nav-link}

A **proposition** is a statement that is known to be true or known to be false. If we don't know if a statement is true or false, then the statement is not a proposition.



## Step-3 {.nav-item}

### {.nav-link}

**(a)**

The statement is a proposition because we know that the statement is true 
$T$ *(as Boston is the capital of Massachusetts)*.



## Step-4 {.nav-item}

### {.nav-link}

Let us summarize how the results were derived.

To determine whether a statement is a proposition, we determined whether the statement was true or false.

If we were able to derive a truth value, then the statement has to be a proposition.
If we were unable to assign a truth value to the statement or the truth value was dependent on some variable, then the statement couldn't be a proposition.



## Results {.nav-item}

### {.nav-link}

*(a) Proposition, $T$*

:::


## Exercise 1b {.tabset .tabset-fade}

### Step-1

In this exercise, we determine whether the given statement represents a proposition.

**What is a proposition?**



### Step-2

A **proposition** is a statement that is known to be true or known to be false. If we don't know if a statement is true or false, then the statement is not a proposition.



### Step-3

(b)

The statement is a proposition because we know that the statement is false $F$ *(as Tallahassee is the capital of Florida)*.



### Step-4

Let us summarize how the results were derived.

To determine whether a statement is a proposition, we determined whether the statement was true or false.

  - If we were able to derive a truth value, then the statement has to be a proposition.
  - If we were unable to assign a truth value to the statement or the truth value was dependent on some variable, then the statement couldn't be a proposition.

<br>

### Results

*(b) Proposition, $F$*


